Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2022-11-07 nakayama/paper.tex
Updates to first part of Formal Proof section

%\author{}
\author{ 
	Andrew Cousino \\ {\ttfamily acousino@ku.edu}
	\and 
	Emily E.\ Witt \\ {\ttfamily witt@ku.edu}	
	\and 
	Perry Alexander \\ {\ttfamily palexand@ku.edu}
}	
	%

@@

%\thanks{Witt acknowledges support from NSF CAREER Award DMS-1945611 and the 2022-23 Ruth I.\, Michler Memorial Prize from the Association for Women in Mathematics.}

\todo[inline]{To do:  Add to/alter author list, title, and/or abstract as necessary}

@@

Our goal is to newly formalize theoretical, rather than computational, commutative algebra in Coq. 
We formally prove \emph{Nakayama's Lemma} \cite{nakayama-1951, azumaya}, an essential result in the field. 
In doing so, we formalize algebraic structures that are fundamental to higher-level algebra, such as 
\emph{local rings} and \emph{(finitely generated) modules over commutative rings}. 
Rather than build upon some of the basic objects from abstract algebra, such as groups and rings, that have been formalized in Coq, e.g., in the  
\emph{Mathematical Components Library} 
\cite{mathcomp}, we start from scratch. \todo{To do: Verify whether MathComp only formalized finite algebraic structures}
 The theory, including the formalization of all algebraic structures, is approximately  $100$ kB and \textcolor{red}{??}\todo{To do:  Add number} lines of code.

@@

\emily[inline]{Emily: It might be more consistent to require a ring to contain $1$ in general, instead of adding this to the axioms of a commutative ring. After thinking for a while, I think it is a good idea to make this change in the code, and then here to match.  Drew, does this sound OK?}

@@

\begin{figure}[t]
\caption{The hierarchy of our algebraic structures}
\includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
\centering
\todo[inline]{To do:  Check source code to verify whether the hierarchy figure should be altered (subobjects, quotients, finitely generated modules, local rings, etc.)}
\end{figure}

@@

We proved that an ideal $I$ of a commutative rings that contains a unit, then it must be the entire ring.  A unit is 
an element $x$ with a multiplicative inverse $x^{-1}$, i.e., the for which the product $x \cdot x^{-1}$ is the multiplicative identity $1$.  By the absorption property, since $x \in I$, we have that $x \cdot x^{-1} = 1 \in I$, and so for every element $r$ of $R$, $r = r \cdot 1$ is also in $I$, i.e., $I=R$.

Next, we formally defined a maximal ideal, a proper ideal that are maximal with respect to inclusion, i.e., if it is strictly contained in another ideal, this larger ideal must be the entire ring. 
Below is the definition in Coq,
which uses \texttt{P} as the predicate for the ideal.

@@

\emily[inline]{Emily:  I'm thinking this discussion could possibly be moved.  I'll think about it a bit more.}

Here we had to include an axiom that in commutative ring, any non-unit \(x\) is
contained in some maximal ideal. This was made into an axiom as the standard
mathematical argument is a proof with potentially infinitely many steps. The
standard argument goes as follows.
\begin{quote}
    Set \(I_{1}\coloneqq (x)\) to be the principal ideal for \(x\), i.e. the
    ideal generated by the single element \(x\). If \(I_{1}\) is not maximal,
    then there exists a strictly larger ideal \(I_{2}\), i.e.
    \(x\in I_{1}\subsetneq I_{2}\). If \(I_{2}\) is not maximal, then there
    exists another strictly larger ideal \(I_{3}\). Continue these arguments
    infinitely many times if necessary in order to get an infinite chain of
    ideals containing \(x\).
    \[x\in I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq R\]
    It is a simple matter to show that \(\bigcup_{k=1}^{\infty} I_{k}\) is also
    an ideal containing in \(x\). So by Zorn's lemma, there exists a maximal
    ideal in \(R\) which contains \(x\).
\end{quote}
Zorn's lemma is equivalent to the axiom of choice. So we had to add at least
one axiom here in order to proceed. By adding the axiom that we did, not only
will we encapsulate an infinite argument, but we will also avoid the need for
the axiom of choice.

Also, we used classical logic to prove that \(1 - x\) is a unit where \(x\) is
a non-unit in any local ring. The proof used the rule \(\neg\neg P\rightarrow
P\) in order to do a proof by contradiction.

\emily[inline]{Emily:  Unfortunately, I think we need to update naming conventions since in mathematics ``vector'' is reserved for an element of a vector space over a field, rather than an element of an arbitrary module, causing confusion.  Similar with ``basis,'' which is especially tricking since two different generating sets of a finitely generated module can have different sizes, but all bases of a vector space have the same cardinality, and an element of a module can typically be written as a scalar combination in multiple ways, while bases are linearly independent.  I've proposed a rewrite of the blue paragraph above it.  Drew, can you check it for accuracy with respect to the code? 
}
\emily[inline]{
We'll also need to update the code.  I think {\tt Vectors} should be something like {\tt ScalarCombinations}. 
}

@@

\subsection{Formally Proving of Nakayama's Lemma}

@@

@@

We needed a lemma before moving on to prove the main theorem which states that
for a finitely generated module \(M\) with a basis
\(b_{1}, b_{2}, \dots, b_{n}\), any element \(x\in I M\) where \(I\) is an
ideal can be written as a linear combination of the basis vectors with
coefficients coming from \(I\). The proof of this lemma is straightforward and
follows the standard, informal mathematical argument which inductively goes
through the vectors needed to generate \(x\) and shows that each vector can be
rewritten in terms of the basis vectors times elements of the ideal \(I\) as
follows.
\begin{align*}
  x & = \sum_{k=1}^{m} u_{k} a_{k} \\
    & = \sum_{k=1}^{m} u_{k} (r_{k1} b_{1} + \cdots + r_{kn} b_{n}) \\
    & = \sum_{j=1}^{n} (u_{1} r_{1j} + \cdots + u_{m} r_{mj}) b_{j}
\end{align*}
Since each \(u_{k}\) comes from the ideal \(I\), the absorbing property of
ideals guarantees that \(u_{k} r_{k j}\) is also contained in \(I\). As ideals
are also closed under addition, it follows that \(u_{1} r_{1j} + \cdots +
u_{m} r_{m j}\) are all elements of \(I\). Thus, \(x\) can be written as a
linear combination of the basis vectors with the coefficients coming from
\(I\).

>>>>

%\author{}
\author{ 
	Andrew Cousino \\ {\ttfamily acousino@ku.edu}
	\and 
	Emily E.\ Witt\thanks{Witt acknowledges support from NSF CAREER Award DMS-1945611 and the 2022-23 Ruth I.\, Michler Memorial Prize from the Association for Women in Mathematics.}
 \\ {\ttfamily witt@ku.edu} \and 
	Perry Alexander \\ {\ttfamily palexand@ku.edu}
}	
	%

@@

\todo[inline]{To do:  Add to/alter author list, funding acknowledgments, title, and/or abstract as needed}

@@

Our goal is to newly formalize theoretical, rather than computational, commutative algebra in Coq. 
We formally prove \emph{Nakayama's Lemma} \cite{nakayama-1951, azumaya}, an essential result in the field. 
In doing so, we formalize algebraic structures that are fundamental to higher-level algebra, such as 
\emph{local rings} and \emph{modules over commutative rings}, and \emph{quotient rings and modules}.  
Rather than build upon some of the basic objects from abstract algebra, such as groups and rings, that have been formalized in Coq, e.g., in the  
\emph{Mathematical Components Library} 
\cite{mathcomp}, we start from scratch. \todo{To do: Verify whether MathComp only formalized finite algebraic structures}
 The theory, including the formalization of all algebraic structures, makes up approximately $100$ kB and 3300 lines of code.

@@

\emily[inline]{Emily: It would be more consistent to require a ring to contain $1$ in general, instead of adding this to the axioms of a commutative ring.  After thinking for a while, I think it is a good idea to make this change in the code, and then here to match.  Drew, does this sound OK?}

@@

\begin{figure}[t]
\caption{The hierarchy of our algebraic structures}
\includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
\centering
\todo[inline]{To do:  Double check source code to verify whether the hierarchy figure should be altered (subobjects, quotients, finitely generated modules, local rings, etc.)}
\end{figure}

@@

\emily[inline]{It would be nice to differentiate between our structures that are typeclasses (e.g., {\tt Ideal}), and other definitions (e.g., {\tt maximal ideal})}
Next, we formally defined a maximal ideal, a proper ideal that are maximal with respect to inclusion, i.e., if it is strictly contained in another ideal, this larger ideal must be the entire ring. 
Below is the definition in Coq,
which uses \texttt{P} as the predicate for the ideal.

@@

\emily[inline]{Emily:  Unfortunately, I think we need to update naming conventions since in mathematics, ``vector'' is reserved for an element of a vector space over a field, rather than an element of an arbitrary module.  So this could cause confusion.  Similar with ``basis,'' which is especially tricking since two different generating sets of a finitely generated module can have different sizes, but all bases of a vector space have the same cardinality, and an element of a module can typically be written as a scalar combination in multiple ways, while bases are linearly independent.  I've proposed a rewrite of the blue paragraph right above it.  Drew, can you double check it for accuracy with respect to the code? 
}
\emily[inline]{
We'll also need to update the code.  I think {\tt vectors} should be something like {\tt scalar-combinations}, and {\tt basis} something like {\tt generating-set}. 
}

@@

\subsection{Building the Formal Proof}

@@

Beyond formalizing the relevant structures from higher algebra, we also formally establish some basic theory. 
For instance, a \emph{unit} of a commutative ring $R$ is an element $x \in R$ with a multiplicative inverse, i.e., an element $x^{-1} \in R$ for which $x \cdot x^{-1}$ is the multiplicative identity $1 \in R$.

We formally proved that an ideal $I$ of a commutative ring $R$ that contains a unit $x$, then $I$ must be the trivial ideal, i.e., the entire ring.  The informal logic is as follows:  By the absorption property, since $x \in I$, we have that $x \cdot x^{-1} = 1 \in I$.  Hence for every element $r$ of $R$, $r = r \cdot 1$ is also in $I$, i.e., $I=R$.

Every non-unit element of a commutative ring is contained in some maximal ideal. 
This fact relies on the Axiom of Choice. 
The following standard informal proof calls upon Zorn's lemma, which is equivalent to the Axiom of Choice, and says that given a partially ordered set $S$, if every chain in $S$ has an upper bound, then $S$ must have at least one maximal element.

\begin{quote}
    Set \(I_{1}\) to be the principal ideal generated by an element \(x\) of a commutative ring $R$, i.e. the
    smallest ideal containing the element $x$, which consists of all $R$-multiples of $x$.

    If \(I_{1}\) is not a maximal ideal,
    then there exists a strictly larger proper ideal \(I_{2}\) of $R$, i.e.
    $x\in I_{1}\subsetneq I_{2} \subsetneq R$.
    Moreover, if  \(I_{2}\) is not maximal, then there
    exists a strictly larger proper ideal $I_{3} \supsetneq I_2$.

    Continuing this process, if it terminates at some step, we have found a maximal ideal, and if not, 
    one obtains an infinite chain of
    ideals containing \(x\).
    \[x\in I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq R\]

    It is straightfoward to show that \(\bigcup_{k=1}^{\infty} I_{k}\) satisfies the definition of an ideal of $R$, 
    and this union certainly contains \(x\). So by Zorn's lemma, since every ascending chain of ideals containing $x$ 
    with respect to inclusion has its union as an upper bound, there exists a maximal
    ideal of \(R\) containing \(x\).
\end{quote}

This argument has potentially infinitely many steps, 
and chose to avoid this issue by including an axiom that in 
 any non-unit \(x\) is
contained in some maximal ideal.

\todo[inline]{Include some relevant code}

We also used classical logic to prove that \(1 - x\) is a unit where \(x\) is
a non-unit in any local ring. The proof used the rule \(\neg\neg P\rightarrow
P\) in order to do a proof by contradiction.

\todo[inline]{Include some relevant code}

@@

We needed a lemma before moving on to prove the main theorem which states that
for a finitely generated $R$-module \(M\) with generating set\emily{Changed ``basis'' to ``generating set'' here}
\(u_{1}, u_{2}, \dots, u_{n}\), any element \(x\in I M\), where \(I\) is an
ideal of $R$, can be written as a linear combination of the basis vectors with
coefficients coming from \(I\). \emily{I'm confused:  Isn't this how we defined $IM$?}
 The proof of this lemma is straightforward and
follows the standard, informal mathematical argument, which inductively goes
through the vectors needed to generate \(x\) and shows that each vector can be
rewritten in terms of the basis vectors times elements of the ideal \(I\) as
follows.
\begin{align*}
  x & = \sum_{k=1}^{m} u_{k} a_{k} \\
    & = \sum_{k=1}^{m} u_{k} (r_{k1} b_{1} + \cdots + r_{kn} b_{n}) \\
    & = \sum_{j=1}^{n} (u_{1} r_{1j} + \cdots + u_{m} r_{mj}) b_{j}
\end{align*}
Since each \(u_{k}\) comes from the ideal \(I\), the absorbing property of
ideals guarantees that \(u_{k} r_{k j}\) is also contained in \(I\). As ideals
are also closed under addition, it follows that \(u_{1} r_{1j} + \cdots +
u_{m} r_{m j}\) are all elements of \(I\). Thus, \(x\) can be written as a
linear combination of the basis vectors with the coefficients coming from
\(I\).
